\documentclass{ucbthesis}
\usepackage[doi=true,url=false,eprint=false,isbn=false,firstinits=true]{biblatex}
\usepackage{amsmath,amsfonts,graphicx,amsthm,pdflscape}

\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}
  
  
 \makeatletter
\newtheorem*{rep@theorem}{\rep@title}
\newcommand{\newreptheorem}[2]{%
\newenvironment{rep#1}[1]{%
 \def\rep@title{#2 \ref{##1}}%
 \begin{rep@theorem}}%
 {\end{rep@theorem}}}
\newcommand{\thickhline}{%
    \noalign {\ifnum 0=`}\fi \hrule height 1.5pt
    \futurelet \reserved@a \@xhline
}
\makeatother

% Double spacing, if you want it.
% \def\dsp{\def\baselinestretch{2.0}\large\normalsize}
% \dsp

% If the Grad. Division insists that the first paragraph of a section
% be indented (like the others), then include this line:
 \usepackage{indentfirst}


\newtheorem{theorem}{Theorem}
\newreptheorem{theorem}{Theorem}
\DeclareMathOperator{\tr}{tr}
\newcommand{\B}[0]{\mathbf}
\newcommand{\BS}[0]{\boldsymbol}
\setcounter{tocdepth}{2}
\setcounter{secnumdepth}{2}

\bibliography{Dissertation}

\hyphenation{Tsah-kadsor Bio-metrika smooth-ed}
%\hyphenation{mar-gin-al-ia}

\begin{document}

% Declarations for Front Matter

\title{Longitudinal Cluster Analysis with Applications to Growth Trajectories}
\author{Brianna Christine Heggeseth}
\degreesemester{Spring}
\degreeyear{2013}
\degree{Doctor of Philosophy}
\chair{Professor Nicholas Jewell}
\othermembers{Assistant Professor Cari Kaufman\\
  Associate Professor Alan Hubbard}
\numberofmembers{3}
\prevdegrees{B.A. (Saint Olaf College) 2008}
\field{Statistics}
\campus{Berkeley}



\maketitle
%\approvalpage
\copyrightpage

\include{Abstract}

\begin{frontmatter}

\begin{dedication}
\null\vfil
\begin{center}
To Paul, my partner in life.\\\vspace{12pt}

\end{center}
\vfil\null
\end{dedication}

\begin{KeepFromToc}
  \tableofcontents
\end{KeepFromToc}

\clearpage
\listoffigures
\clearpage
\listoftables

\begin{acknowledgements}
This dissertation would never have come to fruition without the support of many individuals, and it is with pleasure that I acknowledge them.

I would like to express my utmost gratitude to my advisor, Nicholas Jewell, for giving me the freedom to explore unfamiliar territory and guidance when I needed it the most. He has taught me how to ask good questions and concisely express my ideas. I am indebted to him for helping me establish my career by providing opportunities for collaboration and professional development. Many thanks to Alan Hubbard and Cari Kaufman for their expertise and support as members of my dissertation committee. I want to thank Brenda Eskenazi and Kim Harley for access to data from the CHAMACOS cohort study and their contributed expertise regarding child development. I further thank and acknowledge Bin Yu, Deborah Nolan, and Barbara Abrams for their unending support and encouragement throughout my time at Berkeley. I strive to be the inspirational role model these three women have been to me.

I want to thank all of the faculty in the department for providing me with a strong foundation on which I complete my research and the staff for providing administrative and computing support. 

Thanks to \emph{Statistics in Medicine} for accepting a subset of the material presented in Chapters 1 and 2 for inclusion in the publication titled, `The Impact of Covariance Misspecification in Multivariate Gaussian Mixtures on Estimation and Inference: An Application to Longitudinal Modeling' to appear soon \cite{heggeseth2013}. 

Many fellow graduate students have helped me stay sane through these challenging years. Their support and care helped me overcome stumbling blocks and continue my work. I greatly value their friendship and I deeply appreciate the time we have had together.

Most importantly, none of this would have been possible without the love and patience of my family My parents told me to set my sights high and pushed me to reach my full potential. My friend, partner, and husband, Paul, never gives up on me and brings joy each day to my life.
\end{acknowledgements}

\end{frontmatter}



\pagestyle{headings}

%(Optional) \part{First Part}

\include{Chapter1}
\include{Chapter2}
\include{Chapter3}
\include{Chapter4}
\include{Chapter5}
\include{Chapter6}
\include{Chapter7}


\printbibliography

\appendix
\include{Appendix1} 
\include{Appendix2}


\end{document}
